A Roadmap for\\HEP Software and Computing R\&D\\for the 2020s

### 2018-11-08 CWP/papers/HSF-CWP-2017-01_roadmap/latex/cwp-main.tex
Address review comments
- Add extra citation for HEP SLOC and rephrase
- Refer to appendix for visualisation workshop*s*
- Correct LaTeX formatting error

In order to process the data, the 4 largest LHC experiments have written
more than 20 million lines of program code over the last 15 years. This
has involved contributions from thousands of physicists and many
computing professionals, encompassing a wide range of skills and
abilities. The majority of this code was written for a single
architecture (x86\_64) and with a serial processing model in mind. There
is considerable anxiety in the experiments that much of this software is
not sustainable, with the original authors no longer in the field and
much of the code itself in a poorly maintained state, ill-documented, and
lacking tests. This code, which is largely experiment-specific, manages
the entire experiment data flow, including data acquisition, high-level
triggering, calibration and alignment, simulation, reconstruction (of
both real and simulated data), visualisation, and final data analysis.

@@

Even with the throttling of clock speed to limit power consumption,
power remains a major issue. Low power architectures are in huge demand.
At one level this might challenge the dominance of x86\_64 by simply
replacing it with, for example, AArch64 devices that may achieve lower power
costs for the scale of HEP computing needs than Intel has achieved with its
Xeon architecture. More extreme is an architecture that would see
specialised processing units dedicated to particular tasks, but with
possibly large parts of the device switched off most of the time,
so-called dark silicon~\cite{IntelCSA}.

@@

The workshop that was held as part of the CWP process was felt to be
extremely useful for exchanging knowledge between developers in
different experiments, fostering collaboration and in bringing in ideas
from outside the community. This will now be held as an annual event and
will facilitate work on the common R\&D plan.

@@

To maximise productivity, it is very beneficial to use development tools
that are not HEP-specific. There are many open source projects that are
of similar scale to large experiment software stacks and standard tools
are usually well documented. For source control HEP has generally chosen
to move to \emph{git}~\cite{git},
which is very welcome, as it also brings an
alignment with many open source projects and commercial organisations.
Likewise, \emph{CMake}~\cite{CMake}
is widely used for the builds of software
packages, both within HEP and outside. Packaging many build products
together into a software stack is an area that still requires close
attention with respect to active developments (the HSF has an active
working group here).

>>>>

In order to process the data, the 4 largest LHC experiments have written
tens of millions of lines of program code over the last 15 years
\cite{1742-6596-898-7-072013,Elmer2014}. This
has involved contributions from thousands of physicists and many
computing professionals, encompassing a wide range of skills and
abilities. The majority of this code was written for a single
architecture (x86\_64) and with a serial processing model in mind. There
is considerable anxiety in the experiments that much of this software is
not sustainable, with the original authors no longer in the field and
much of the code itself in a poorly maintained state, ill-documented, and
lacking tests. This code, which is largely experiment-specific, manages
the entire experiment data flow, including data acquisition, high-level
triggering, calibration and alignment, simulation, reconstruction (of
both real and simulated data), visualisation, and final data analysis.

@@

Even with the throttling of clock speed to limit power consumption,
power remains a major issue. Low power architectures are in huge demand.
At one level this might challenge the dominance of x86\_64 by simply
replacing it with, for example, AArch64 devices that may achieve lower power
costs for the scale of HEP computing needs than Intel has achieved with its
Xeon architecture~\cite{1742-6596-608-1-012001}. More extreme is an architecture that would see
specialised processing units dedicated to particular tasks, but with
possibly large parts of the device switched off most of the time,
so-called dark silicon~\cite{martin2014,IntelCSA}.

@@

The workshops that were held as part of the CWP process
(\emph{HSF Visualization Workshop}, see Appendix \ref{appendix-a---list-of-workshops})
were felt to be
extremely useful for exchanging knowledge between developers in
different experiments, fostering collaboration and in bringing in ideas
from outside the community. These will now be held as an annual events and
will facilitate work on the common R\&D plan.

@@

To maximise productivity, it is very beneficial to use development tools
that are not HEP-specific. There are many open source projects that are
of similar scale to large experiment software stacks and standard tools
are usually well documented. For source control HEP has generally chosen
to move to \emph{git}~\cite{git},
which is very welcome, as it also brings an
alignment with many open source projects and commercial organisations.
A major benefit that has come with this technical choice is the use
of social coding sites, such as \emph{GitHub}~\cite{GitHub}
and \emph{GitLab}~\cite{GitLab}, where
code sharing and code review are far superior compared to previous
solutions.
Likewise, \emph{CMake}~\cite{CMake}
is widely used for the builds of software
packages, both within HEP and outside. Packaging many build products
together into a software stack is an area that still requires close
attention with respect to active developments (the HSF has an active
working group here).
